## Setup

Install the required dependencies if not already installed:
```{r}
# The simulator
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
if (!requireNamespace("dependentsimr", quietly = TRUE))
    remotes::install_github("tgbrooks/dependentsimr")

# DESeq2 - optional dependency only needed if doing RNA-seq data sets
# Needed for type = "DESeq2"
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("DESeq2", quietly = TRUE))
    BiocManager::install("DESeq2")

# Two more dependencies that are used in some simulation modes
# Needed for method = "corpcor"
if (!requireNamespace("corpcor", quietly)) 
    install.packages("corpcor")

# Needed for method = "spiked Wishart"
if (!requireNamespace("sparsesvd"))
    install.packages("sparsesvd")

# Extras needed for this vignette
if (!requireNamespace("tidyverse", quietly = TRUE))
    install.packages("tidyverse")
```

Load the libraries:
```{r}
library(dependentsimr)
library(dplyr)
library(tibble)
library(DESeq2)
```

Set a seed for reproducibility.
```{r}
set.seed(0)
```

## Load data

Simulations with `dependentsimr` start with a real data set that should be mimicked.
This data can be almost any shape, including continuous, or discrete.
Here, we will simulate RNA-seq, which has discrete counts that we handle with a special "DESeq2" model.
Other data types can use normal, Poisson, or empirical (which uses only the values present in the data set) types, or even a combination of these.

Read in metadata containing sample names, sample IDs and sex.
```{r}
metadata <- read.delim("/home/jianingy/dependent_sim/dependent_sim_paper/processed/GSE151923_sample_metadata.txt", sep="\t")
```


Read in the data containing read counts of genes for each sample.
```{r}
read_counts <- read.delim("/home/jianingy/dependent_sim/dependent_sim_paper/data/GSE151923_metaReadCount_ensembl.txt.gz", sep="\t")
```


Set rownames to be the gene IDs.
```{r}
row.names(read_counts) <- read_counts[,1]

read_counts <- read_counts[,6:ncol(read_counts)]
```

Drop NAs - the data file is missing part of the very last row, giving NAs.
```{r}
read_counts <- read_counts[!apply(is.na(read_counts), 1, any),]
```

Drop a sample that has very low read counts.
```{r}
read_depth <- apply(read_counts, 2, sum)

READ_DEPTH_CUTOFF = 10e6
read_counts <- read_counts[,read_depth > READ_DEPTH_CUTOFF]
```

Separate by sex. We will only use data from male mice in this example.
```{r}
sex <- vapply(colnames(read_counts),
    function(id) {
        return(metadata[metadata["sample_id"] == id, "sex"])
    }, "")

male_read_counts <- read_counts[,sex == 'male']
```

## Get Dependence Structure

The first step is to compute a 'random structure' for the data, which encodes how the variables (genes, in this case) in the data set are dependent upon each other.
This computes a covariance matrix which captures the dependence between genes, however, that matrix is not the same as the usual sample covariance matrix.
It also computes the shape (such as mean and variance) of each variable individually.

There are three implemented methods ("pca", "spiked Wishart", and "corpcor") with different characteristics.
We demonstrate here each.

First, the "pca" method takes a `rank` and includes only gene-gene dependence up to that many dimensions.
Typically small values (2 or 3, for example) would be used for `rank`, similar to when performing PCA, unless a large number of input samples are available.
Note that we put our `read_counts` data into a `list()`.
This is done since `get_random_structure` also supports multiple data sets of different 'modes' (for example, proteomics and transcriptomics done on the same samples).
We only have one 'mode' that we give the arbitrary name of `data`.
```{r}

rs_pca <- get_random_structure(list(data=read_counts), method="pca", rank=2, types="DESeq2")
```

Second, the spiked Wishart method also takes a `rank` value but uses it differently.
Typically, this will be larger, up to the total number of samples minus one, which is what we use here.
For very large sample counts, `rank` might need to be capped well below the number of samples for computational efficiency reasons.
```{r}
rs_wishart <- get_random_structure(list(data=male_read_counts), rank=11, method="spiked Wishart", types="DESeq2")
```

Third, a method based off the `corpcor` package is implemented.
It takes no parameters but often underestimates the amount of gene-gene dependence in the data.
```{r}
rs_corpcor <- get_random_structure(list(data=male_read_counts), method="corpcor", types="DESeq2")
```

Lastly, we could generate data without gene-gene dependence for comparison by removing the dependence from any of our random structures.
```{r}
rs_indep <- remove_dependence(rs_pca)
```

## Generate Data

Real RNA-seq data sets typically have variation between samples in the total number of reads per sample.
Here, we just copy the total number of reads from the real data set to get realistic values.
A `library_size` of 1 corresponds to the average read depth of the data used to generate the random structure.
This step is only needed when generating RNA-seq data with `type = "DESeq2"`.

```{r}
actual_library_sizes <- male_read_counts |> apply(2, sum)

N_SAMPLES <- 6
library_sizes <- sample(actual_library_sizes / mean(actual_library_sizes), size=N_SAMPLES, replace=TRUE)
```

### Simulate Control Data

Now we generate some 'control' samples, which have the same expression levels (as well as gene-gene dependence) as was estimated from the original data set.
We'll use `rs_pca` that we generated above for example, but any of the methods could be used.
```{r}
control_sim_data <- draw_from_multivariate_corr(rs_pca, n_samples=N_SAMPLES, size_factors=library_sizes)$data
```
Note that we used `$data` here because that was the name we provided to `get_random_structure` and it was the only mode of data that we had.
If multiple modes were used to generate the random structure, then we `draw_from_multivariate_corr` returns a list with each of them.

### Simulate Differentially Expressed Data

To generate a set of 'treatment' samples which differ from the 'control' samples, we will have to modify the random structure to reflect the desired expression values of the treatment group.
We do this by just choosing some genes at random and picking a fold-change for each at random.

```{r}
# fraction of all genes to modify in treatment
FRAC_DE <- 0.05
# Min and max value for the log fold change (LFC) for each DE gene chosen
# non-DE genes will have a LFC of 0
LFC_MIN <- 0.2
LFC_MAX <- 2.0
N_DE <- floor(FRAC_DE * nrow(control_sim_data))

# Pick which genes are DE
de_genes <- sample(nrow(male_sim), N_DE)

# Pick the LFCs of each of those genes
de_lfc <- runif(N_DE, LFC_MIN, LFC_MAX) * sample(c(-1, 1), size=N_DE, replace=TRUE)
```

Now, we create a new random structure object and modify it to have the desired expression values.
Information about the marginal distribution of each gene (e.g., mean and variance but not dependence on other genes) is stored in `rs$marginals` for a random structure `rs`.
The specifics depend upon the `type` used in `get_random_structure`, but for `DESeq2`, the `q` value determines the mean expression, so we change that here.

```{r}
rs_treatment <- rs_pca
rs_treatment$marginals$data$q[de_genes] <- 2^(de_lfc) * rs_treatment$marginals$data$q[de_genes]
```

Lastly, we generate the data for the 'treatment' samples.
```{r}
treatment_sim_data <- draw_from_multivariate_corr(rs_treatment, n_samples=N_SAMPLES, size_factors=library_sizes)$data
```

## Plot the generated data

To demonstrate the data, we show a PCA plot of the control and treatment groups
```{r}
sim_data <- cbind(control_sim_data, treatment_sim_data)
pca <- prcomp(t(sim_data), center=TRUE, scale.=TRUE, rank.=2)
pca_points <- predict(t(sim_data))
pca_data <- tibble(x=pca_points[,1], y=pca_points[,2]) |>
    mutate(group = c(rep("control", N_SAMPLES), rep("treatment", N_SAMPLES)))
ggplot(pca_data, aes(x, y, color=group))
```
