## Setup

Install the required packages, and choose a simulation method (one of "indep", "pca", "corpcor" and "wishart").

```{r}
library(dependentsimr)
library(dplyr)
library(tibble)
library(DESeq2)

method <- 'pca'
```

Read in metadata containing sample names, sample IDs and sex.
```{r}
metadata <- read.delim("/home/jianingy/dependent_sim/dependent_sim_paper/processed/GSE151923_sample_metadata.txt", sep="\t")
```


Read in the data containing read counts of genes for each sample.
```{r}
read_counts <- read.delim("/home/jianingy/dependent_sim/dependent_sim_paper/data/GSE151923_metaReadCount_ensembl.txt.gz", sep="\t")
```


Set rownames to be the gene IDs.
```{r}
row.names(read_counts) <- read_counts[,1]

read_counts <- read_counts[,6:ncol(read_counts)]
```

Drop NAs - the data file is missing part of the very last row, giving NAs.
```{r}
read_counts <- read_counts[!apply(is.na(read_counts), 1, any),]
```

Drop a sample that has very low read counts.
```{r}
read_depth <- apply(read_counts, 2, sum)

READ_DEPTH_CUTOFF = 10e6
read_counts <- read_counts[,read_depth > READ_DEPTH_CUTOFF]
```

Separate by sex. We will only use data from male mice in this example.
```{r}
sex <- vapply(colnames(read_counts),
    function(id) {
        return(metadata[metadata["sample_id"] == id, "sex"])
    }, "")

male_read_counts <- read_counts[,sex == 'male']
```

## Get Dependence Structure

For the given simulation method, get the dependence structure from the real data set.

```{r}

# Produce simulated data
if (method == "pca") {
    set.seed(1)
    # Simulate with the PCA method
    rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, types="DESeq2")
} else if (method == "corpcor") {
    set.seed(2)
    # Simulate with the corpcor method
    rs <- get_random_structure(list(data=male_read_counts), method="corpcor", types="DESeq2")
} else if (method == "wishart") {
    set.seed(3)
    # Simulate with the spiked Wishart method
    rs <- get_random_structure(list(data=male_read_counts), rank=11, method="spiked Wishart", types="DESeq2")
} else if (method == "indep") {
    set.seed(4)
    # Simulate without any dependence
    dep_rs <- get_random_structure(list(data=male_read_counts), method="pca", rank=2, type="DESeq2")
    rs <- remove_dependence(dep_rs)
}
```


Get the read depth of samples in the real data set, and generate library sizes of the simulated data sets (5 data sets each with 20 samples).
```{r}
actual_library_sizes <- male_read_counts |> apply(2, sum)

N_SAMPLES <- 100
library_sizes <- sample(actual_library_sizes / mean(actual_library_sizes), size=N_SAMPLES, replace=TRUE)
```

## Simulate Control Data

Generate and save data+metadata (true values) from a specific random structure
```{r}
generate <- function(rs, name) {
  sim <- draw_from_multivariate_corr(rs, n_samples=N_SAMPLES, size_factors=library_sizes)$data
  colnames(sim) <- paste("sample", 1:ncol(sim), sep='')

sim |>
    data.frame() |>
    rownames_to_column("ENSEMBL_ID") |>
    write.table(file=paste0("/home/jianingy/dependent_sim/dependent_sim_paper/notebook_data/", name, ".txt"), sep="\t", row.names=FALSE)

  # metadata
  truth <- data.frame(
    gene_id = rownames(sim),
    mean = rs$marginals$data$q,
    disperion = rs$marginals$data$dispersion
  )
  truth[is.na(truth$mean), 2] <- 0 # NA mean is 0
  
  return(sim)
}

male_sim <- generate(rs, name=paste0("Mouse.Cortex.Male.",method,".Control"))
```

## Simulate Differentially Expressed Data

Randomly select some genes to be DE.
```{r}
FRAC_DE <- 0.05
LFC_MIN <- 0.2
LFC_MAX <- 2.0
N_DE <- floor(FRAC_DE * nrow(male_sim))

de_genes <- sample(nrow(male_sim), N_DE)
```


Simulate log fold changes for the case data sets.
```{r}
de_lfc <- runif(N_DE, LFC_MIN, LFC_MAX) * sample(c(-1, 1), size=N_DE, replace=TRUE)
```


Generate Cases data sets.
```{r}
if (method == "indep") {
    male_de_rs <- dep_rs
    male_de_rs$marginals$data$q[de_genes] <- 2^(de_lfc) * male_de_rs$marginals$data$q[de_genes]
    
    male_indep_de_rs <- remove_dependence(male_de_rs)
    male_indep_de_sim <- generate(male_indep_de_rs, name="Mouse.Cortex.Male.indep.Case")
} else {
    male_de_rs <- rs
    male_de_rs$marginals$data$q[de_genes] <- 2^(de_lfc) * male_de_rs$marginals$data$q[de_genes]
    
    male_de_sim <- generate(male_de_rs, name=paste0("Mouse.Cortex.Male.",method,".Case"))
}
```

